---
title: "Summary Statistics"
subtitle: "Using R to Create Basic Summary Tables"
output: html_output
---

## Opening the R packages needed for this workshop

```{r chunk1}
# if needed, install packages through the Tools menu: tidyverse, haven, table1
# load packages needed for this workshop
library(package = "tidyverse")
library(package = "haven")
library(package = "table1")
```


## Importing the data

```{r chunk2}
# 2015 Behavioral Risk Factor Surveillance Survey data
temp <- tempfile(fileext = ".zip")
download.file(url  = "https://www.cdc.gov/brfss/annual_data/2015/files/LLCP2015XPT.zip", destfile = temp)
BRFSS_2015 <- read_xpt(file = temp)
```

## Selecting variables of interest

* Search the codebook to learn how these variables were measured:
    + https://www.cdc.gov/brfss/annual_data/2015/pdf/codebook15_llcp.pdf

```{r chunk3}
# select sex and transgender identity variables
# and alcohol consumption variables
BRFSS_2015_small <- BRFSS_2015 %>%
   select(SEX, TRNSGNDR, `_RFDRHV5`, `_DRNKWEK`)
```

## Examine the variables of interest

```{r chunk4}
# summarize small data set
summary(object = BRFSS_2015_small)
```

## Recoding the sex variable

```{r chunk5}
BRFSS_2015_small <- BRFSS_2015 %>%
  select(SEX, TRNSGNDR, `_RFDRHV5`, `_DRNKWEK`) %>% 
  mutate(SEX = recode_factor(SEX,
                             '2' = 'male',
                             '1' = 'female'))
  
```

# Checking the work so far

```{r chunk6}
# summarize the data frame
summary(object = BRFSS_2015_small)
```

## Recoding the trans variable

```{r chunk7}
BRFSS_2015_small <- BRFSS_2015 %>%
  select(SEX, TRNSGNDR, `_RFDRHV5`, `_DRNKWEK`) %>%  
  mutate(SEX = recode_factor(SEX,
                             '1' = 'male',
                             '2' = 'female')) %>% 
  mutate(TRNSGNDR = recode_factor(TRNSGNDR,
                                  '1' = 'male-to-female',
                                  '2' = 'female-to-male',
                                  '3' = 'gender non-conforming',
                                  '4' = 'not transgender',
                                  '7' = NA_character_,
                                  '9' = NA_character_)) 
  
```

# Checking the work so far

```{r chunk8}
# summarize the data frame
summary(object = BRFSS_2015_small)
```

# You try it!

* Add to the code to add labels to the heavy drinking variable

```{r Chunk9}
# add on to the previous data management
BRFSS_2015_small <- BRFSS_2015 %>%
  select(SEX, TRNSGNDR, `_RFDRHV5`, `_DRNKWEK`) %>%  
  mutate(SEX = recode_factor(SEX,
                             '1' = 'male',
                             '2' = 'female')) %>% 
  mutate(TRNSGNDR = recode_factor(TRNSGNDR,
                                  '1' = 'male-to-female',
                                  '2' = 'female-to-male',
                                  '3' = 'gender non-conforming',
                                  '4' = 'not transgender',
                                  '7' = NA_character_,
                                  '9' = NA_character_))


```

# Checking the work so far

```{r chunk10}
# summarize the data frame
summary(object = BRFSS_2015_small)
```


## Recoding other variables

```{r chunk11}
BRFSS_2015_small <- BRFSS_2015 %>%
  select(SEX, TRNSGNDR, `_RFDRHV5`, `_DRNKWEK`) %>%  
  mutate(SEX = recode_factor(SEX,
                             '1' = 'male',
                             '2' = 'female')) %>% 
  mutate(TRNSGNDR = recode_factor(TRNSGNDR,
                                  '1' = 'male-to-female',
                                  '2' = 'female-to-male',
                                  '3' = 'gender non-conforming',
                                  '4' = 'not transgender',
                                  '7' = NA_character_,
                                  '9' = NA_character_)) %>% 
  mutate(`_RFDRHV5` = recode_factor(`_RFDRHV5`,
                                    '1' = 'no',
                                    '2' = 'yes',
                                    '9' = NA_character_)) %>% 
  mutate(`_DRNKWEK` = as.numeric(`_DRNKWEK`)/100) %>% 
  mutate(`_DRNKWEK` = na_if(`_DRNKWEK`, 999)) %>% 
  rename(sex = SEX) %>% 
  rename(genderIdentity = TRNSGNDR) %>% 
  rename(heavyDrink = `_RFDRHV5`) %>% 
  rename(drinksWeek = `_DRNKWEK`) %>% 
  drop_na()
  
```

# Checking the work so far

```{r chunk12}
# summarize the data set
summary(object = BRFSS_2015_small)
```

## Table of all variables

```{r chunk13}
# basic table with all variables
table1(~ sex + genderIdentity + heavyDrink + drinksWeek, 
       data = BRFSS_2015_small)
```

# Add labels to the table

```{r chunk14}
# labels for each variable
label(BRFSS_2015_small$sex) = "Participant sex"
label(BRFSS_2015_small$genderIdentity) = "Participant gender identity"
label(BRFSS_2015_small$heavyDrink) = "Participant is heavy drinker"
label(BRFSS_2015_small$drinksWeek) = "Number of alcohol drinks per week"

# table of all variables with labels
table1(~ sex + genderIdentity + heavyDrink + drinksWeek, 
       data = BRFSS_2015_small)
```

## Stratified table

```{r chunk15}
# stratified table
table1(~ genderIdentity + heavyDrink + drinksWeek | sex, 
       data = BRFSS_2015_small)
```

## Available descriptive statistics for table1

* Good resource document for the `table1` package: https://cran.r-project.org/web/packages/table1/table1.pdf

* Available descriptive statistics (* default): 
    + N* -- number of non-missing observations 
    + NMISS -- number missing values
    + MEAN* -- mean
    + SD* -- standard deviation
    + MIN* -- minimum value
    + MEDIAN* -- median or middle value
    + MAX* -- maximum value
    + IQR -- interquartile range 
    + CV -- coefficient of variation
    + GMEAN -- geometric mean
    + GSD -- geometric standard deviation
    + GCV -- geometric coefficient of variation
    + FREQ* -- frequency 
    + PCT* -- percentage

## Distribution of continuous variable

```{r chunk16}
BRFSS_2015_small %>% 
  ggplot(aes(x = drinksWeek)) +
  geom_histogram() +
  facet_wrap(facets = vars(sex))
```

## Modifying the table

```{r chunk17}
# specifying continuous variable stats
table1(~ genderIdentity + heavyDrink + drinksWeek | sex, 
       render.continuous = 'Median (iqr)',
       data = BRFSS_2015_small)
```

## Fix the label to identify metrics

```{r chunk18}
# add median and iqr to label
label(BRFSS_2015_small$drinksWeek) = "Number of alcohol drinks per week (median, iqr)"

# specifying continuous variable stats
table1(~ genderIdentity + heavyDrink + drinksWeek | sex, 
       render.continuous = 'Median (iqr)',
       data = BRFSS_2015_small)
```

## You try it! 

* Change the table1 code so that it:
    + stratifies by heavyDrink rather than by sex
    + prints the median with min and max for the continuous variables

```{r chunk19}
# table stratified by heavyDrink
table1(~ genderIdentity + heavyDrink + drinksWeek | sex, 
       data = BRFSS_2015_small)
```


## Other base R descriptive statistics code

```{r chunk20}
mean(BRFSS_2015_small$drinksWeek, na.rm = TRUE)

median(BRFSS_2015_small$drinksWeek, na.rm = TRUE)

sd(BRFSS_2015_small$drinksWeek, na.rm = TRUE)

IQR(BRFSS_2015_small$drinksWeek, na.rm = TRUE)

table(BRFSS_2015_small$sex)

table(BRFSS_2015_small$sex, BRFSS_2015_small$heavyDrink)

prop.table(table(BRFSS_2015_small$sex))
```

## descr package for more summary statistics

* descr -- this package has functions for computing group means, crosstabs, and other frequently used descriptive statistics 

    + https://cran.r-project.org/web/packages/descr/descr.pdf 
    
```{r chunk21}
# install descr first
# compare means function example
library(package = "descr")
compmeans(x = BRFSS_2015_small$drinksWeek, 
          f = BRFSS_2015_small$sex,
          plot = FALSE)
```

    
## Hmisc package for more summary statistics

* Hmisc -- this package has a great `describe` function that gives a detailed list of descriptive statistics for every variable in a data frame 

    + https://cran.r-project.org/web/packages/Hmisc/Hmisc.pdf

```{r chunk22}
# install Hmisc first
# describe function
library(package = "Hmisc")
describe(BRFSS_2015_small)
```


## The end

* Email me at harrisj@wustl.edu

* Connect with me on Twitter @jenineharris

* Check out my book, Statistics with R: Solving Problems Using Real World Data: https://us.sagepub.com/en-us/nam/statistics-with-r/book253567 at Sage (Coupon: SAGE30)

