---
title: "Activity"
author: "Kim Johnson"
date: 
output: html_document
---

## Instructions. For this activity, we are going to use the BRFSS 2017 dataset (a 10% random sample) to analyze height and weight data overall and by sex and age of the participant.


## Open library for importing a .csv file.

```{r}
library(readr) 
```

## Import BRFSS data and assign it the name BRFSS.

```{r}
BRFSS<-read_csv("https://raw.githubusercontent.com/jenineharris/apha-2021/main/BRFSS2017_10percent_v2.csv")
```

## Determine the number of rows (number of observations) and columns/variables.

```{r}
dim(BRFSS)
```

## Get the variable names.

```{r}
colnames(BRFSS)
```

## Delete the first column and check that it's removed.

```{r}
BRFSS <- subset(BRFSS, select = -c(...1) )
colnames(BRFSS)
```

## Move rowID to be first and check that it is first.

```{r}
BRFSS<-BRFSS[, c(20, 1:19)]
View(BRFSS)
```

## Rename X_AGE5YR to AGE5YR.

```{r}
names(BRFSS)[3]<-c("AGE5YR") 
colnames(BRFSS)
```

## Get the mean weight and mean height.

```{r}
mean(BRFSS$wtkg, na.rm = TRUE)
mean(BRFSS$ht_meters, na.rm = TRUE)
```

## Check the values for sex using the table function.

```{r}
table(BRFSS$SEX)
```
# Create a factor variable called SEX_f for sex using the ifelse function with 0=Male and 1=Female. In the original dataset the 1=Male, 2=Female, and 9=Refused.  Assign NAs to 9.

```{r}
BRFSS$SEX_f<-ifelse(BRFSS$SEX==1, 0, ifelse(BRFSS$SEX==2, 1, NA))
BRFSS$SEX_f<-factor(BRFSS$SEX_f, levels=c(0,1), 
              labels = c("Male", "Female")) 
```

# Run the table function on your new variable to see if the categorization worked correctly. Not you can add the argument useNA="always" to see the n of NAs.

```{r}
table(BRFSS$SEX_f, useNA = "always")
```

# Subset males and get the mean height and weight.

```{r}
males<-filter(BRFSS, SEX_f=='Male')
mean(males$ht_meters, na.rm=TRUE) 
mean(males$wtkg, na.rm=TRUE) 
```

# Subset females and get the mean height and weight.

```{r}
females<-filter(BRFSS, SEX_f=='Female')
mean(females$ht_meters, na.rm=TRUE) 
mean(females$wtkg, na.rm=TRUE) 
```

# Merge in the BMI variable by rowID from the file called "BRFSS2017_BMI.csv

```{r}
BRFSSbmi<-read_csv("https://raw.githubusercontent.com/jenineharris/apha-2021/main/BRFSS2017_BMI.csv")
BRFSSall<-merge(BRFSSbmi, BRFSS, by="rowID")
```
# Find the mean, min, and max bmi

```{r}
summary(BRFSSall$bmi,  na.rm=TRUE)

```

# Categorize bmi according to CDC categories at https://www.cdc.gov/healthyweight/assessing/bmi/adult_bmi/index.html

```{r}
BRFSSall$bmi_cat<-ifelse(BRFSSall$bmi<18.5, 0, 
                         ifelse(BRFSSall$bmi>=18.5 & BRFSSall$bmi<25, 1, 
                                ifelse(BRFSSall$bmi>=25 & BRFSSall$bmi<30, 2,
                                       ifelse(BRFSSall$bmi>=30, 3, NA))))
```

# Create a bmi category factor variable and label the categories according to the CDC labels

```{r}
BRFSSall$bmi_cat<-factor(BRFSSall$bmi_cat, levels=c(0:3), 
              labels = c("Underweight", "Healthy weight", "Overweight", "Obesity")) 
```

# Run the table function on your bmi categorical variable to check categorization

```{r}
table(BRFSSall$bmi_cat)
```
# Useful function not discussed to check the variable for accuracy
- by(numerical variable name, categorical variable name, summary)

```{r}
#by(BRFSSall$bmi, BRFSSall$bmi_cat,  summary)
```

