---
title: "Activity"
author: "Kim Johnson"
date: 
output: html_document
---

## Instructions. For this activity, we are going to use the BRFSS 2017 dataset (a 10% random sample) to analyze height and weight data overall and by sex and age of the participant.


## Open library for importing a .csv file.

```{r}

```

## Import BRFSS data and assign it the name BRFSS.

```{r}

```

## Determine the number of rows (number of observations) and columns/variables.

```{r}

```

## Get the variable names.

```{r}

```

## Delete the first column and check that it's removed.

```{r}

```

## Move rowID to be first and check that it is first.

```{r}

```

## Rename X_AGE5YR to AGE5YR.

```{r}

```

## Get the mean weight and mean height.

```{r}

```

## Check the values for sex using the table function.

```{r}

```
# Create a factor variable called SEX_f for sex using the ifelse function with 0=Male and 1=Female. In the original dataset the 1=Male, 2=Female, and 9=Refused.  Assign NAs to 9.

```{r}

```

# Run the table function on your new variable to see if the categorization worked correctly. Not you can add the argument useNA="always" to see the n of NAs.

```{r}

```

# Subset males and get the mean height and weight.

```{r}

```

# Subset females and get the mean height and weight.

```{r}

```

# Merge in the BMI variable by rowID from the file called "BRFSS2017_BMI.csv

```{r}

```
# Find the mean, min, and max bmi

```{r}

```

# Categorize bmi according to CDC categories at https://www.cdc.gov/healthyweight/assessing/bmi/adult_bmi/index.html

```{r}
BRFSSall$bmi_cat<-ifelse(BRFSSall$bmi<18.5, 0, 
                         ifelse(BRFSSall$bmi>=18.5 & BRFSSall$bmi<25, 1, 
                                ifelse(BRFSSall$bmi>=25 & BRFSSall$bmi<30, 2,
                                       ifelse(BRFSSall$bmi>=30, 3, NA))))
```

# Create a bmi category factor variable and label the categories according to the CDC labels

```{r}

```

# Run the table function on your bmi categorical variable to check categorization

```{r}

```
# Useful function not discussed to check the variable for accuracy
- by(numerical variable name, categorical variable name, summary)

```{r}

```

