---
title: "You must be plotting something"
subtitle: "Data visualization in ggplot2"
output: html_output
---

## How to follow along and participate

```{r chunk1}
library(package = "tidyverse")
```

## Examine the full data set

```{r chunk2}
starwars
```

## Recode a few variables in the data set

```{r chunk3}
# change data types & subset data set to remove lists
# recode hermaphroditic to intersex modern terminology
# recode species to fewer categories
starwarsRec <- starwars %>% 
  select(-films, -vehicles, -starships) %>% 
  mutate(sex = recode_factor(sex,
                             'hermaphroditic' = 'intersex')) %>% 
  mutate(species = if_else(species %in% c('Human', 'Droid'), species, 'Other')) %>% 
  mutate(eye_color = if_else(eye_color %in% c('brown', 'black', 'dark'), 
                             'dark', 
                             if_else(eye_color %in% c('blue', 'hazel', 'blue-gray', 'red,blue'),
                                     'light',
                                                      if_else(eye_color == 'unknown',
                                                              NA_character_,
                                                              'bright')))) %>% 
  mutate_if(is.character, as.factor)
  
```

## Check the recoding

```{r chunk4}
# summarize updated data set
summary(object = starwarsRec)
```

## Basic bar chart

```{r chunk5}
starwarsRec %>%
  ggplot(aes(x = gender)) +
  geom_bar() 
```

## Fancy bar chart

```{r chunk6}
starwarsRec %>%
  drop_na(gender) %>% 
  ggplot(aes(x = gender)) +
  geom_bar(fill = "#7463AC") +
  labs(x = "Character Gender",
       y = "Frequency",
       title = "Gender of Star Wars Characters") +
  theme_minimal()
```

## You try it!

Make a bar chart of the Species of Star Wars Characters in your R Markdown file.

Copy, paste, and edit the fancy bar chart code from chunk6 to: 

* Change the gender variable to the species variable

* Change the color of the bars by replacing #7463AC the name of your favorite color (e.g., orange) or using a color found on the https://colorbrewer2.org/ website

* Add labels for the axes and a title inside the quote marks

* Try another theme, you can find them in the help documents by searching "complete themes" or here: https://ggplot2.tidyverse.org/reference/ggtheme.html

When you have completed the changes, click the little green arrow on the top right of the code chunk to run the code.


```{r chunk7}
# star wars species bar chart


```


## The end

* Email me at harrisj@wustl.edu

* Connect with me on Twitter @jenineharris

* Check out my book, Statistics with R: Solving Problems Using Real World Data: https://us.sagepub.com/en-us/nam/statistics-with-r/book253567 at Sage (Coupon: SAGE30)

