---
title: 'APHA Learning Institute: Introduction to R Statistical Software for Public
  Health'
author: "Kim Johnson"
date:
output: html_document
---

## Assign the number 3 to the object x

```{r, chunk1}
x<-3
```

## Create data frame

```{r, chunk2}
df<-data.frame(x)
```

## Making a vector and printing them

```{r, chunk3}
x<-c(1,2,3,4,5)
y<-c(1:5)
x
y
```

## Creating and printing a list

```{r, chunk4}
mylist<-c("c", "a", 3, 4, "dataframe")
mylist
```

## Creating a simple function and checking it

```{r, chunk5}
myfun<-function(v){
  mean(v)
}

myfun(v=x)
myfun(v=y)

#Check with mean function that already exists
mean(x)
mean(y)
```

## Installing packages and opening libraries

```{r, chunk6}
# install.packages("haven") 
# install.packages("readr") 
# install.packages("dplyr")
# install.packages("openxlsx")

# for SPSS (read_sav()), stata (read_dta()), SAS(read_sas()) and xpt (read_xpt()) files
library(haven) 

# for reading csv file (read_csv())
library(readr) 

# for data management
library(dplyr) 

#for exporting excel files
library(openxlsx) 
```

## Importing data: read in csv file

```{r, chunk7}
starbucks_csv <-
 read_csv(
   "https://raw.githubusercontent.com/jenineharris/apha-2021/main/starbucks_csv.csv"
   )
```

## Importing data: read in stata file

```{r, chunk8}
starbucks_stata <-
  read_dta(
    "https://github.com/jenineharris/apha-2021/blob/main/starbucks_stata.dta?raw=true"
  )
```

## Importing data: read in tab file

```{r, chunk9}
starbucks_tab <-
  read.delim(
    "https://raw.githubusercontent.com/jenineharris/apha-2021/main/starbuckstab_tab.txt"
  )
```

## Importing data: read in SAS XPT file

```{r, chunk10}
starbucks_xpt <-
  read_xpt(
    "https://github.com/jenineharris/apha-2021/blob/main/starbuck_xpt.xpt?raw=true"
  ) 
```

## Importing data: read SPSS file

```{r, chunk11}
starbucks_spss <-
  read_sav(
    "https://github.com/jenineharris/apha-2021/blob/main/starbucks_spss.sav?raw=true")
```

## Importing data: from your computer when it is located in the same folder as your code

```{r, chunk12}
starbucks_spss.same<-read_sav("starbucks_spss.sav")
```

## Importing data: from your computer using file path

```{r, chunk13}
#starbucks_spss.different<-read_sav("your path")
```

## Importing data: using R studio's Import Dataset pull down menu

```{r, chunk14}
#insert code here
```

## Viewing the dataset

```{r, chunk15}
View(starbucks_xpt)
```

## Characterizing the dataset (we will use starbucks_xpt)

```{r, chunk16}
dim(starbucks_xpt) 
```

## Getting a summary of the types of variables using sapply

```{r, chunk17}
sapply(starbucks_xpt, typeof)
```

## Getting a summary of the type of variable using class

```{r, chunk18}
class(starbucks_xpt$DRINK)
```

## Getting column names to rename variables

```{r, chunk19}
colnames(starbucks_xpt)
```

## Using the names function 

```{r, chunk20}
#renames variables in order of appearance
names(starbucks_xpt)[1:7]<-c("drink", "category", "calories", "fat (g)", 
                             "carb. (g)", "fiber (g)", "protein (g)") 

colnames(starbucks_xpt) 
```

## Renaming only one variable/column

```{r, chunk21}
#renames variables in order of appearance
names(starbucks_xpt)[2]<-c("cat") 
colnames(starbucks_xpt)
```

## Using the rename function from the dplyr package to rename specific variables

```{r, chunk22}
starbucks_xpt<-rename(starbucks_xpt, c("CALORIES"="calories"))
```

## Converting the character variable CALORIES to numeric
```{r, chunk23, echo=TRUE, warning=FALSE}

#first check class again
class(starbucks_xpt$CALORIES)

#convert calories variable to numeric using the as.numeric function for stats
starbucks_xpt$calories_n<-as.numeric(starbucks_xpt$CALORIES) 

#check the class of the new numeric calorie variable
class(starbucks_xpt$calories_n)
```

## Getting simple statistics

```{r, chunk24, echo=TRUE, warning=FALSE}
#get summary stats
summary(starbucks_xpt$calories_n) 

#get mean
mean(starbucks_xpt$calories_n, na.rm=TRUE) 

#get median
median(starbucks_xpt$calories_n, na.rm=TRUE) 

#get sd
sd(starbucks_xpt$calories_n, na.rm=TRUE) 

#get variance
var(starbucks_xpt$calories_n, na.rm=TRUE) 

#get quantile
quantile(starbucks_xpt$calories_n, na.rm=TRUE) 
```

## Deleting columns 

```{r, chunk 25}
#using the subset function
starbucks_xpt2 <- subset(starbucks_xpt, select = -c(CALORIES) )
colnames(starbucks_xpt2)

#you can also do it the reverse by keeping all but the variables you want to drop
starbucks_xpt3 = subset(starbucks_xpt, select = c(drink, cat, `fat (g)`, `carb. (g)`, `fiber (g)`, `protein (g)`))
colnames(starbucks_xpt3)

#rename cat back to category
starbucks_xpt<-rename(starbucks_xpt, c("category"="cat"))
colnames(starbucks_xpt)
```

## Deleting rows 

```{r, chunk26}
starbucks_xpt4<-na.omit(starbucks_xpt)
```

## Deleting a specific row

```{r, chunk27}
starbucks_xpt4<-starbucks_xpt4[-1,]
```

## Categorize/creating a factor variable 

```{r, chunk28}
#at or below the median
starbucks_xpt$calories_med<-ifelse(starbucks_xpt$calories_n<=140, 0, ifelse(starbucks_xpt$calories_n>140, 1, NA))
```

## Check the type of variable

```{r, chunk29}
class(starbucks_xpt$calories_med) 
```

## Use the **as.factor** function to change from a numerical variable to a factor variable and label levels 

```{r, chunk30}
starbucks_xpt$calories_med<-factor(starbucks_xpt$calories_med, levels=c(0,1), 
              labels = c("Less than or equal to the median", "Above the median")) 
```

## Make sure it's a factor

```{r, chunk31}
class(starbucks_xpt$calories_med) 
```

## Get the n in each level and check that our categorization is correct using the table function

```{r, chunk32}
table(starbucks_xpt$calories_med) 
```

## Subsetting: using the base **which** function

```{r, chunk33}
espresso<-starbucks_xpt[which(starbucks_xpt$category=='Starbucks Espresso Beverages'),] 

# check that subsetting worked by checking number of espresso drinks in parent dataframe
table(starbucks_xpt$category)

# check that subsetting worked by checking number of espresso drinks in child dataframe
dim(espresso) 

# calculate mean number of calories in espresso drinks, removing 'NAs' first. 
# Reminder: if you don't remove na's, the mean function will not work
mean(espresso$calories_n, na.rm=TRUE) 

#another way to see the mean number of calories in espresso drinks is using the summary function
summary(espresso$calories_n) 

# and another way if you don't want to create a new dataframe
mean(starbucks_xpt[which(starbucks_xpt$category=='Starbucks Espresso Beverages'),]$calories_n, na.rm=TRUE)
```

## Subsetting: using the **filter** function from the dplyr package

```{r, chunk34}
espresso<-filter(starbucks_xpt, category=='Starbucks Espresso Beverages')
mean(espresso$calories_n, na.rm=TRUE) 
```

## Changing a value of a variable

```{r, chunk35}
espresso$calories_n[espresso$drink=='Brown Sugar Shortbread Latte']<-300
```

## Changing a value of a variable using the **ifelse** function

```{r, chunk36}
espresso$calories_n<-ifelse(espresso$drink=='Brown Sugar Shortbread Latte', 300, espresso$calories_n)
```

## Creating datasets to use as examples for **cbind**, **rbind**, and **merge** functions

```{r, chunk37}
employee <- c('Trayan Klavdiya','Wasswa Tadesse','Jianhong Dong', "Mary Smith")
salary <- c(210000, 234000, 268000, 240000)
startdate <- as.Date(c('2010-11-1','2008-3-25','2007-3-14', '2009-10-28'))
```

## Combining each vector into three columns

```{r, chunk38}
employ.data <- data.frame(employee, salary, startdate, stringsAsFactors = FALSE)
class(employ.data)

# this makes a matrix (which can be useful for math functions) instead of a 
# dataframe but we can convert it to a data frame using the as.data.frame function
employ.data2<-cbind(employee, salary, startdate) 

class(employ.data2)
employ.data2<-as.data.frame(employ.data2)
class(employ.data2)
```

## Adding a new column to our new dataset using the **cbind** function

```{r, chunk39}
coffee_pref<-c("dark", "light", "pumpkin spice", "likes tea")

# the stringsAsFactors argument prevents R from turning character variables 
# into factor variables
employ.data<-cbind(employ.data, coffee_pref, stringsAsFactors = FALSE) 
class(employ.data)
```

## Adding a new row using the **rbind** function

```{r, chunk40}
new_employee<-c('Martha Kuskowski', 240500, '2020-11-01', 'pumpkin spice')
employ.data<-rbind(employ.data, new_employee)
View(employ.data)
```

## Creating another dataframe 

```{r, chunk41}
employee <- c('Trayan Klavdiya','Wasswa Tadesse','Jianhong Dong', "Mary Smith", 
              'Martha Kuskowski')
termdate<-as.Date(c('2021-11-1','2021-11-1','2021-11-1', '2021-11-1', 
                    '2021-11-1'))
employ.data2<-data.frame(employee, termdate)

#merge employ.data and employ.data2 by employee name
all.employee.data<-merge(employ.data, employ.data2, by="employee")
```

## Assigning ID numbers

```{r, chunk42}
all.employee.data$ID <- seq.int(nrow(all.employee.data))
View(employ.data)
```

## Reordering columns 

```{r, chunk43}
all.employee.data<-all.employee.data[, c(6,1,2,3,4,5)]
View(all.employee.data)
```

## Exporting a dataset as an excel file or a csv file using the **write.xlsx** and **write.csv** functions

```{r, chunk44}
# write an excel file
write.xlsx(all.employee.data, "all.employee.data.xlsx", overwrite=TRUE)

# write a csv file
write.csv(all.employee.data, "all.employee.data.csv")
```


# Contact information

- kijohnson@wustl.edu